---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(gridExtra)

faithful <- tibble(faithful)
`   ``


```{r theme_dark_blue}
# gotten from https://github.com/koundy/ggplot_theme_Publication

theme_dark_blue <- function(base_size=14, base_family="sans") {
   library(grid)
   library(ggthemes)
   (theme_foundation(base_size=base_size, base_family=base_family)
      + theme(plot.title = element_text(face = "bold", colour = '#ffffb3',
                                        size = rel(1.2), hjust = 0.5, margin = margin(0,0,20,0)),
              text = element_text(),
              panel.background = element_rect(colour = NA, fill = '#282C33'),
              plot.background = element_rect(colour = NA, fill = '#282C33'),
              panel.border = element_rect(colour = NA),
              axis.title = element_text(face = "bold",size = rel(1), colour = 'white'),
              axis.title.y = element_text(angle=90,vjust =2),
              axis.title.x = element_text(vjust = -0.2),
              axis.text = element_text(colour = 'grey70'), 
              axis.line.x = element_line(colour="grey70"),
              axis.line.y = element_line(colour="grey70"),
              axis.ticks = element_line(colour="grey70"),
              panel.grid.major = element_line(colour="#343840"),
              panel.grid.minor = element_blank(),
              legend.background = element_rect(fill ='#282C33'),
              legend.text = element_text(color = 'white'),
              legend.key = element_rect(colour = NA, fill = '#282C33'),
              legend.position = "bottom",
              legend.direction = "horizontal",
              legend.box = "vetical",
              legend.key.size= unit(0.5, "cm"),
              #legend.margin = unit(0, "cm"),
              legend.title = element_text(face="italic", colour = 'white'),
              plot.margin=unit(c(10,5,5,5),"mm"),
              strip.background=element_rect(colour="#2D3A4C",fill="#2D3A4C"),
              strip.text = element_text(face="bold", colour = 'white')
      ))
}
```

This is the typical dataset that is used to demonostrate the need for mixture models. A good example why it's important to visualise data before using it can be seen here.

```{r}
faithful %>%
  ggplot(aes(x = eruptions, y = waiting)) +
  geom_point() +
  theme_dark_blue()
```

It looks like there are two different distributions or populations in this set. 

```{r}
faithful %>%
  ggplot(aes(x = waiting)) +
  geom_density(adjust = 1L) +
  theme_dark_blue()
```




